---
title: "Stable Isotope Analysis"
author: "Masumbuko Semba and Nyamisi Peter"
date: "`r format(Sys.time(), 'On %d %B, %Y at %H:%M:%OS3')`"
output: html_document
editor_options: 
  chunk_output_type: console
link-citations: yes
# csl: havard.csl
bibliography: isotopes.bib
---

# Introduction
In many case we have data from different sampling locations. or different types of individuals (male/female) or different tissues (muscle and fin) and we compare between these groups. For our data we can compare betwen the sampling locations because the data were collected at eight different sites. We first create a group of habitat and simmr wants these variable to be as integer, we therefore convert the sites into integer.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, comment=""}
setwd("E:/Data Manipulation/Kim/isotopes/")
```

```{r, comment="", message=FALSE, warning=FALSE}
require(tidyverse)
require(simmr)
require(DT)
require(cowplot)

```

# Loading the Data
The first step involved loading the data into R workspace. @readxl developed *readxl* package that can read Microsoft Excel Spreadsheet directly from the local machine. The sheet number three that contains isotopic data was imported from the local machine into the workspace of [R](https://www.r-project.org/)---a free and open source software for statistical, computation, analysis and graphics [@r]
```{r, comment=""}
## read the file
kim = readxl::read_excel("Clean_Data.xlsx", sheet = 1)%>%na.omit()

```

# Tidy the data
```{r}
## tidy the data
kim = kim%>%
    separate(Sample, c("Sample", "number"), sep = 1, remove = T)%>%
    mutate(site.grp = Site, sample = Sample,
         site.grp = replace(site.grp, site.grp == "Kunduchi", 1),
         site.grp = replace(site.grp, site.grp == "Mbegani", 2),
         sample = replace(sample, sample == "F", 1),
         sample = replace(sample, sample == "T", 2))%>%
    select(-number, Species = Spp, Area = Site,  Site = site.grp)
## convert the character into integer
kim$Site = as.integer(kim$Site)
kim$sample = as.integer(kim$sample)


```

```{r, comment=""}
knitr::kable(kim, caption = "Table 1: Overview of Isotopes")
```

```{r, comment=""}
knitr::kable(kim%>%group_by(Area,Species, Sample)%>%count()%>%arrange(desc(n))%>%mutate(percentage = n/84*100), digits = 2, caption = "Table 2: Summary statistics of the six common species")
```

# Step 2: Getting the data into simmr
Once we have the dataset in the workspace, we have to tidy in the format that *simmr* package understand [@tidyr] . This is because simmr requires at minimum 3 input objects; the *mixtures*, the *source* means, and the source *standard deviations* as matrix [@simmr].

>Optionally, you can also add correction data (also called trophic enrichment factors, TEFs) represented again as means and standard deviations, and concentration dependence values. 

```{r, comment=""}
mix = kim%>%select(1,2)%>%as.matrix()

```
The created mix object contains the stable isotopic data for the consumers.  The matrix function was used to turn this into a matrix (a rectangle of numbers) with 2 columns. The first column contains the data for isotope 1, and the second the data for isotope 2. 



```{r, comment=""}

mean.data = kim%>%group_by(Species)%>%
  summarise(d13C = mean(d13C), d15N = mean(d15N))%>%
  ungroup()

s_means = mean.data[,2:3]%>%as.matrix()


sd.data = kim%>%group_by(Species)%>%
  summarise(d13C = sd(d13C), d15N = sd(d15N))%>%
  ungroup()

s_sds = sd.data[,2:3]%>%as.matrix()

s_names = mean.data$Species


```

## compare between sites and species
```{r}
sites = kim$Site

simmr.in.group.sites = simmr_load(mixtures = mix,
           source_names = s_names,
           source_means = s_means,
           source_sds = s_sds, 
           group = sites)

```


```{r}

```


## compare between tissues and species
```{r}
tissue = kim$sample

simmr.in.group.sample = simmr_load(mixtures = mix,
           source_names = s_names,
           source_means = s_means,
           source_sds = s_sds, 
           group = tissue)

```

